% UHCUtrechtHaskellCompiler-AU.tex
\begin{hcarentry}[section,updated]{UHC, Utrecht Haskell Compiler}
\label{uhc}
\label{ehc}
\report{Atze Dijkstra}%11/10
\status{active development}
\participants{many others%Jeroen Fokker,
% Doaitse Swierstra,
% Arie Middelkoop,
% Luc\'ilia Camar\~{a}o de Figueiredo,
% Carlos Camar\~{a}o de Figueiredo,
% Andres L\"oh, Jos\'e~Pedro Magalh\~{a}es,
% Vincent van Oostrum, Clemens Grabmayer, Jan Rochel,
% Tom Lokhorst, Jeroen Leeuwestein, Atze van der Ploeg, Paul van der Ende, Calin Juravle, Levin Fritz
}
\makeheader

\paragraph{What is new?}
UHC is the Utrecht Haskell Compiler, supporting almost all Haskell98 features and most of Haskell2010, plus
experimental extensions.
Recently version 1.1.0 was released, featuring generic deriving \cref{sec:genericDeriving},
a new configurable garbage collector, initial support for building with UHC via cabal, and many bug fixes.

We plan the next release to offer a Javascript backend.
Furthermore we hope to add optimizations to eliminate some of the obvious inefficiencies.
As part of this work the intent is also to better integrate the work done on whole program analysis.

\paragraph{UHC Blog}
Recently a UHC blog has been started.
The intent is to scribble about the internals of UHC and issues arising out of the implementation.

\paragraph{What do we currently do and/or has recently been completed?}
As part of the UHC project, the following (student) projects and other activities are underway (in arbitrary order):
\begin{itemize}
\item Jeroen Bransen (PhD): ``Incremental Global Analysis'' (starting up).
\item Jan Rochel (PhD): ``Realising Optimal Sharing'', based on work by Vincent van Oostrum and Clemens Grabmayer.
\item Arie Middelkoop (PhD): type system formalization and automatic generation from type rules.
\item Tom Lokhorst: type based static analyses (recently completed, available with proper configuration).
\item Jeroen Leeuwestein: incrementalization of whole program analysis.
\item Atze van der Ploeg: lazy closures (recently completed).
\item Paul van der Ende: garbage collection \& LLVM (recently completed, available with proper configuration).
\item Jeroen Fokker: GRIN backend, whole program analysis.
\item C\u{a}lin Juravle: base libraries (completed upto Haskell98, integrated).
\item Levin Fritz: base libraries for Java backend (completed, integrated).
\item Andres L\"oh: Cabal support (completed initial support, integrated).
\item Jos\'e~Pedro Magalh\~{a}es: generic deriving (\cref{sec:genericDeriving}, completed, integrated, presented at Haskell Symposium).
\item Doaitse Swierstra: parser combinator library.
\item Atze Dijkstra: overall architecture, type system, bytecode interpreter + java + javascript backend, garbage collector.
\end{itemize}

\paragraph{Background}

UHC actually is a series of compilers of which the last is UHC, plus
infrastructure for facilitating experimentation and extension.
The distinguishing features for dealing with the complexity of the compiler and for experimentation are
(1) its stepwise organisation as a series of increasingly more complex standalone compilers,
the use of DSL and tools for its (2) aspectwise organisation (called Shuffle) and
(3) tree-oriented programming (Attribute Grammars, by way of the
Utrecht University Attribute Grammar (UUAG) system~\cref{uuag}.
%
%For more information, see the references provided.

\FurtherReading
\begin{compactitem}
\item UHC Homepage:
\url{http://www.cs.uu.nl/wiki/UHC/WebHome}

\item UHC Blog:
\url{http://utrechthaskellcompiler.wordpress.com}

\item Attribute grammar system:
\url{http://www.cs.uu.nl/wiki/HUT/AttributeGrammarSystem}

\item Parser combinators:
\url{http://www.cs.uu.nl/wiki/HUT/ParserCombinators}

\item Shuffle:
\url{http://www.cs.uu.nl/wiki/Ehc/Shuffle}

\item Ruler:
\url{http://www.cs.uu.nl/wiki/Ehc/Ruler}
\end{compactitem}
\end{hcarentry}
