\begin{hcarentry}{EHC, `Essential Haskell' Compiler}
\label{ehc}
\report{Atze Dijkstra}
\status{active development}
\participants{Atze Dijkstra, Doaitse Swierstra}
\entry{nearly unchanged, pinged}% done, 06.05.2005
\makeheader

The purpose of the EHC project is to provide a description a Haskell compiler
which is as understandable as possible so it can be used for education as well
as research.

For its description an Attribute Grammer system (AG) is used
as well as other formalisms allowing compact notation like parser combinators.
For the description of type rules,
and the generation of an AG implementation for those type rules,
we recently started using the Ruler system (included in the EHC project).

The EHC project also tackles other issues:
\begin{itemize}
\item
   In order to avoid overwhelming the innocent reader,
   the description of the compiler is organised as a series of
   increasingly complex steps.
   Each step corresponds to a Haskell subset which itself is an extension
   of the previous step.
   The first step starts with the essentials, namely typed lambda
   calculus.

\item
   Each step corresponds to an actual, that is, an executable compiler.
   Each of these compilers is a compiler in its own right so
   experimenting can be done in isolation of additional complexity
   introduced in later steps.

\item
   The description of the compiler uses code fragments which are
   retrieved from the source code of the compilers.
   In this way the description and source code are kept synchronized.
\end{itemize}

Currently EHC already incorporates more advanced features like
higher-ranked polymorphism, partial type signatures, class system,
explicit passing of implicit parameters (i.e. class instances),
extensible records, kind polymorphism.

Part of the description of the series of EH compilers is available as a PhD thesis,
which incorporates previously published material on the EHC project.

The compiler is used for small student projects as well as larger
experiments such as the incorporation of an Attribute Grammar system.

We also hope to provide a Haskell frontend dealing with all Haskell
syntactic sugar omitted from EHC.

\FurtherReading
\begin{compactitem}
\item Homepage:

\url{http://www.cs.uu.nl/groups/ST/Ehc/WebHome}

\item Attribute grammar system:

\url{http://www.cs.uu.nl/groups/ST/twiki/bin/view/Center/AttributeGrammarSystem}

\item Parser combinators:

\url{http://www.cs.uu.nl/groups/ST/Software/UU_Parsing/}
\end{compactitem}  
\end{hcarentry}
